Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-21 pnas/Towards.tex
modified discussion  + 1-D figure in the App

@@

In the simplified mathematical model we examined, we showed it is possible to estimate a signal without detecting its appearances.  
Our strategy is to compute autocorrelation functions of the micrographs and to relate these statistics to the unknown signal's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.

@@

Looking towards applying the framework to encompass all important features of real cryo-EM experiments, our work implies that it might be possible to reconstruct small molecules, particularly, molecules that are too small to be detected in micrographs. 
In pursuing this research direction, our goal is to significantly increase the range of molecules to which cryo-EM can be successfully applied.
We recognize that significant challenges lay ahead for the implementation of the proposed approach to 3-D reconstruction directly from the micrographs. We discuss a few now.

@@

Beyond data acquisition and computational challenges, there are modeling issues to consider.
In contrast to the simplifying assumptions we have made above, the noise might be colored; the viewing directions of the particles may be distributed non-uniformly; there may be conformational heterogeneity; particles generally do not satisfy our separation condition; and micrographs undergo a contrast transfer function which we have omitted. All of these aspects can be handled with the same general strategy: establish a forward model relating the expected autocorrelations of the micrographs to the target volume(s) and all parameters necessary to model the above effects. For instance, for colored noise, the forward model may involve multiple parameters to capture the power spectrum of the noise instead of the single parameter $\sigma^2$. The separation condition could be relaxed by assuming a Poisson process of particle distribution in the micrograph, in which case the forward model will involve the Poisson parameter.
%To address this point, it may be necessary to investigate better noise models and to incorporate power spectrum estimation methods. % and to extend the autocorrelation analysis accordingly. 
%Second, we assume that any two particles are sufficiently separated, and we use this assumption to derive algebraic relations between autocorrelations of the micrographs and autocorrelations of the 3-D volume. 
%We believe that this constraint could be replaced by a probabilistic condition on the average number of projections per micrograph.  
%Such a condition can be met in practice by careful experimental design. 
%
%We also note several aspects of SPR experiments that were neglected in this paper, including contrast transfer function (CTF), non-uniformly distributed viewing directions and other contaminants in the micrographs, such as ice crystals.
All these aspects must be taken into account so the method can be applied on experimental data. We hope to take care of these issues in future research.

@@

For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.

>>>>

\begin{bibunit}
%
%\TODO{In RELION 3 paper, see these references, relevant for our claims about streaming mode: ``Other programs tackled individual steps in the single-particle workflow, like [...] on-the-fly processing of images while they are being acquired at the microscope [de la Rosa-Trevin et al., 2016, Biyani et al., 2017].''}

@@

In the simplified mathematical model above, we showed it is possible to estimate a signal without detecting its appearances.  
Our strategy is to compute autocorrelations of the micrographs and to relate these statistics to the unknown signal's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.

@@

Looking toward applying the framework to encompass all important features of real cryo-EM experiments, our work implies that it might be possible to reconstruct small molecules, particularly, molecules that are too small to be detected in micrographs. 
In pursuing this research direction, our goal is to significantly increase the range of molecules to which cryo-EM can be successfully applied.
We recognize that significant challenges lay ahead for the implementation of the proposed approach to \revised{high-resolution} 3-D reconstruction directly from the micrographs. We discuss a few now.

@@

\revised{To reach high-resolution reconstruction,} beyond data acquisition and computational challenges, there are modeling issues  to consider.
In contrast to the simplifying assumptions we have made above, the noise might be colored; the viewing directions of the particles may be distributed non-uniformly; there may be conformational heterogeneity; particles generally do not satisfy our separation condition; and micrographs undergo a contrast transfer function which we have omitted. All of these aspects can be handled with the same general strategy: establish a forward model relating the expected autocorrelations of the micrographs to the target volume(s) and all parameters necessary to model the above effects. 
For instance, for colored noise, the forward model may involve multiple parameters to capture the power spectrum of the noise instead of the single parameter $\sigma^2$. 
\revised{Similarly, instead of the separation condition, we can model the spacing between the projections using a parameterized pair-correlation function. Such a function models the distribution of distances between neighboring projections. The observed autocorrelations depend linearly on these parameters, which would be estimated as part of the inverse problem.
}
%The separation condition could be relaxed by assuming a Poisson process of particle distribution in the micrograph, in which case the forward model will involve the Poisson parameter.
%To address this point, it may be necessary to investigate better noise models and to incorporate power spectrum estimation methods. % and to extend the autocorrelation analysis accordingly. 
%Second, we assume that any two particles are sufficiently separated, and we use this assumption to derive algebraic relations between autocorrelations of the micrographs and autocorrelations of the 3-D volume. 
%We believe that this constraint could be replaced by a probabilistic condition on the average number of projections per micrograph.  
%Such a condition can be met in practice by careful experimental design. 
%
%We also note several aspects of SPR experiments that were neglected in this paper, including contrast transfer function (CTF), non-uniformly distributed viewing directions and other contaminants in the micrographs, such as ice crystals.
All these aspects must be taken into account so the method can be applied on experimental data. We hope to take care of these issues in future research.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{example_obs_1D}
	\caption{\label{fig:1D_obs_example} An example for a 1-D noisy measurement for the experiment in Figure~\ref{fig:1D_example}. The signal occurrences appear in blue, and the noisy data in green.
	}
\end{figure}

For the 1-D experiment depicted in Figure~\ref{fig:1D_example}, we used a signal of length $L=21$ and generated an observation $y$ of length $N = 10(2L-1)M$, where $M$ is the number of repetitions. 
The observation was generated by randomly selecting placements in $y$, one at a time with an accept/reject rule based on the separation condition and locations picked so far until reaching the desired number of signal occurrences. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting SNR of $y$ is about $1/175$.
\revised{Figure~\ref{fig:1D_obs_example} shows an example of the observed data.}
